---
title: "Vignette: An introduction to prettyGraphics"
author: "Edward Lavender^[Centre for Research into Ecological and Environmental Modelling, The Observatory, University of St Andrews, St Andrews, Scotland, KY16 9LZ and Scottish Oceans Institute, East Sands, University of St Andrews, St Andrews, Scotland, KY16 8LB, el72@st-andrews.ac.uk]"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introducing_prettyGraphics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<style>
body{
  width: 100%;
  height: 100%;
  font-family: TimesNewRoman;
  font-size: 12pt;
  text-align: justify;
  }
  p{
  margin-bottom: 1.2em;
  }
</style>

# Introduction

`prettyGraphics` is an R package designed to make the production of plots and data exploration easier, more flexible and prettier, with base R. The creation of this package was inspired by the development of another package, `Tools4ETS`, which provides tools for ecological timeseries in R. Consequently, `prettyGraphics` has been particularly motivated by the requirements of datasets with continuous responses and explanatory variables (e.g., depth and time, in relation to covariates). The package is based on a number of functions which act as 'building blocks'; these help to define the initial arguments of a plot and then add elements to a plot sequentially. The package includes some integrative functions, which link these functions together to produce pretty plots in a variety of cases (e.g. a more flexible implementation of `graphics::plot()` with pretty axes, prettier histograms, density plots, timeseries, model residuals etc.). Notable functionality of `prettyGraphics` includes the following: 

* **The definition of pretty axes**. The definition of 'pretty' axes (i.e., axes with intelligible tick mark labels that are positioned in appropriate, adjoining positions, rather than as an approximate box around a plot) underlies publication-quality plots but these are not implemented automatically by base R. `pretty_axis()` is a very flexible function which makes this possible. A few other functions, such as `sci_notation()`, which can be used to add labels with scientific notation to a plot (i.e. $a \times 10^b$ rather than $ae^b$), and `add_lagging_point_zero()`, which brings all numbers up to the same number of decimal places, complement this function. `add_grid_rect_xy()` can be used to add a rectangular grid at user-defined positions (e.g. coinciding with axis labels), to aid interpretation  
* **Colouring lines by a covariate**. In base R, it's easy to add lines to a plot to illustrate a relationship. However, it is a more long-winded process to add a line that is coloured by the values of a third variable, followed by a highly customisable colour bar legend, to illustrate the relationship between the response and more than one explanatory variable. This essential part of data exploration is implemented by `add_lines()` and `add_colour_bar()`. 
* **Shading**. Adding shading to a plot is another way to illustrate possible relationships, especially between a response/explanatory variable and a factor. For example, in plots of depth timeseries, it is often useful to add shading delineating day/night periods. Adding shading to a plot is made easy with `add_shading_bar()`. 
* **Statistical summaries**. The definition and addition of statistical summarises to a plot can also help to reveal signal from noise for continuous data. For example, it might be helpful to calculate the mean value, the minimum, the maximum, the 89 % quantiles or another statistic in a sequence of bins $n$ units in width (e.g. $10 cm$, $0.5 ms^{-2}$, $12 hours$, etc.). This is facilitated by `summmarise_in_bins()` and `add_lines()`. 
* **Model predictions**. The addition of model predictions (e.g. fitted values ± confidence intervals) on top of observations is critical step in the evaluation of models (e.g. to understand the extent to which they explain observed patterns). This is facilitated by `list_CIs()` and `add_model_predictions()`. 
* **Integrative functions**. `prettyGraphics` provides several functions which use the power of building block functions to create prettier plots, including histograms, density plots, residual plots and a number of others. For example, `pretty_plot()` functions like `graphics::plot()` but uses `pretty_axis()` to define pretty axes and is more flexible. `pretty_plot()` can incorporate multiple plotting functions (not just `graphics::plot()`) but some functions (e.g. `graphics::hist()`) necessitate a specialised plotting approach which `prettyGraphics` may provide (e.g., `pretty_hist()`). 

The aim of this vignette is to introduce some of these functions, to highlight their flexibility and their use, and to show how they can be strung together to produce prettier plots designed to shed light on interesting questions. For each function, much more detail is provided in the help files. 

# Package set up

## Installation and load

First, install and load `prettyGraphics`:
```{r}
# install.packages("prettyGraphics")
library(prettyGraphics)
```

This vignette also uses functions from the following packages, which you may need to install before loading and attaching: 
```{r, echo = TRUE, results = "hide"}
library(magrittr) # pipe, %>%
library(pryr)     # active binding, %<a-%
```

## R prerequisites

`prettyGraphics` and this vignette use a few tricks with which it is helpful to be familiar before proceeding. 

### lists and `do.call()`

The use of lists and `do.call()` is a backbone of `prettyGraphics`. Lists are very useful objects in R which can store all sorts of objects. Lists contain elements, which may be named, that contain these objects. For example, in the following list, the first element is named `a` and it contains the number 10. The second element is unnamed and contains 0; the third element is unnamed and contains a dataframe; and the fourth element is an unamed list which contains the number 16:
```{r}
l <- list(a = 10, 0, data.frame(x = 1, y = 2), list(16))
utils::str(l)
```

One particularly useful feature of lists is that they can be used to hold function arguments. Specifically, a named list of arguments can be passed to a function via `do.call()` for the that function to evaluate. In the following example, I generate a collection of numbers and then I use the `pretty()` function to define a pretty sequence across the range of these numbers, by specifying the `x` and `n` arguments of `pretty()` in a named list:
```{r}
# generate numbers 
x <- stats::runif(10, 0, 10)
# do.call implementation 
pretty_args <- list(x = x, n = 5)
names(pretty_args); utils::str(pretty_args)
p1 <- do.call(pretty, pretty_args)
# standard implementation
p2 <- pretty(x, n = 5)
# do.call() and standard implementation are idential: 
p1; p2; identical(p1, p2)
```

`prettyGraphics` makes use of this functionality. For example, there is a function, `pretty_axis()` which can make pretty axes. This includes an argument, `pretty`, which takes in a list of arguments that are passed to `pretty()` or `lubridate::pretty_dates()` to make pretty axes. Here is a canned example in which I add pretty axes to side 1 and 2 of a plot using `pretty_axis()`. Each axis has approximately five pretty breaks, which the function defines by default based on the data to be plotted (`x` and `y` here). In this case, since `x` and `y` are numeric, the named list of arguments is passed to `pretty()` within the function to define the pretty sequence. `pretty_axzis()` then does some more work to define axis placement and then adds the axes to the plot. 

```{r, fig.height = 3, fig.width = 3, fig.align = "center"}
x <- runif(10, 0, 10)
y <- runif(10, 0, 10)
plot(x, y, xlim = c(0, 10), ylim = c(0, 10), axes = FALSE)
pretty_axis(side = 1:2, x = list(x, y), pretty = list(n = 5), add = TRUE, return_list = FALSE)
```

Under-the-hood, `pretty_axis()` loops over any nested lists, which means each axis can be controlled independently via a nested list. For example, to specify the x and y axes with different numbers of pretty breaks, the following code could be used:
```{r, fig.height = 3, fig.width = 3, fig.align = "center"}
plot(x, y, xlim = c(0, 10), ylim = c(0, 10), axes = FALSE)
pretty_axis(side = 1:2, x = list(x, y), pretty = list(list(n = 5), list(n = 10)), add = TRUE, return_list = FALSE)
```

There are many more examples of the power and flexibility of `pretty_axis()` in the help file (use `?pretty_axis`).

### Active binding and `%<a-%`

This vignette also uses a trick called active binding to minimise code repetition. This is implemented via the `pryr` package. With active binding, the user can save code in an object. When the object is executed, the code that is bound to that object is then executed. Consider the following example. First, we'll make a plot and actively bind this to an object. 

```{r}
# Create active binding; note that no object will show at this stage. 
active_binding_example %<a-% plot(x, y)
```

Then, after executing the object, the code is evaluated and the plot becomes visible:
```{r, fig.height = 3, fig.width = 3, fig.align = "center"}
active_binding_example
```

## Example dataset

To illustrate some of the functionality of `prettyGraphics`, I use an example dataset from the `Tools4ETS` package which is included in `prettyGraphics`. This is a dataset which contains a sample of a depth and temperature observations at defined timestamps collected from two flapper skate, _Dipturus intermedius_, off the West Coast of Scotland by Scottish Natural Heritage and Marine Scotland Science. We'll load these data and focus on one individual for simplicity. For visualisation purposes, we'll also define a new column that contains depth expressed as a negative number:
```{r}
df <- dat_flapper[which(dat_flapper$id == "A"), ]
df$dn <- df$depth*-1
utils::head(df, 3)
utils::str(df)
```

Although we will use these data as an example, it is important to emphasise that`prettyGraphics`'s functions work for many continuous response and explanatory covariates; they are not restricted to timeseries (though they are particularly useful for some timeseries data.) A few functions are also designed to work with discrete data. 

# `prettyGraphics` building blocks 

## Pretty axes: `pretty_axis()`

The first step towards building a publication-quality plot is the definition of suitable axis limits and the addition of suitable axes in appropriate positions. Consider the following example, in which the default plot (produced by `graphics::plot()`) is contrasted with one created using `pretty_axis()`. Note that I choose to plot the x axis on the $3^{rd}$ side in this and following examples so that we can visualise this axis as the 'sea surface', from which we can visualise animal movement into shallower or deeper water.

```{r, fig.width = 7, fig.height = 10}
#### Set plotting window 
pp <- par(mfrow = c(2, 1))

#### Default plot of (negated) depth against time 
plot(df$timestamp, df$dn, 
     type = "l", xlab = "Time", ylab = "Depth")

#### Prettier plot
# First use pretty_axis to define a list of outputs which include axis limits
axis_ls <- pretty_axis(side = c(3, 2), 
                       x = list(df$timestamp, df$dn), 
                       axis = list(las = TRUE),
                       pretty = list(n = 5), 
                       add = FALSE)
# We can use the limits defined by pretty_axis() to create the plot 
plot1 %<a-% plot(df$timestamp, df$dn, 
                 type = "l", 
                 xlab = "Time", ylab = "Depth",
                 axes = FALSE, 
                 xlim = axis_ls[[1]]$lim, ylim = axis_ls[[2]]$lim,
                 ); plot1
# The list of outputs can be passed back to pretty_axis() to add pretty_axes()
# Note the automatic pretty labels and adjoined placing:
add_pretty_axis %<a-% pretty_axis(axis_ls = axis_ls, add = TRUE, return_list = FALSE)
add_pretty_axis
par(pp)

```

`pretty_axis()` has lots of customisation options which are exemplified in the help file (use `?pretty_axis`). In this case, the integrative function, `pretty_plot()`, could be used here to reproduce an identical plot without first calling `pretty_axis()`. This is demonstrated at the end of the vignette. 

## Add lines: `add_lines()`

We now have the outline of a plot. However, it might be helpful to consider not only how depth changes through time, but whether this change is related to temperature. To do this, one option is to colour the depth timeseries by temperature: 

```{r fig.width = 7, fig.height = 5}
#### Define plotting window 
pp <- par(oma = c(1, 1, 1, 4))

#### Plot the graph with pretty limits, hiding the line 
plot2 %<a-% plot(df$timestamp, df$dn, 
                 type = "n",
                 xlab = "Time", ylab = "Depth",
                 axes = FALSE, 
                 xlim = axis_ls[[1]]$lim, ylim = axis_ls[[2]]$lim,
                 ); plot2

#### Use add_lines() to add a line coloured by temperature. 
# The function outputs a list of arguments, which we'll save because we can use them to create
# ... a colour bar later (e.g. this includes pretty axis labels controlled by arguments to pretty_axis_args). 
dat4colbar <- 
  add_lines(x = df$timestamp, 
            y1 = df$dn, 
            y2 = df$temp, 
            f = grDevices::colorRampPalette(c("red", "blue")), 
            pretty_axis_args = list(pretty = list(n = 10), axis = list(las = TRUE)))

#### Use add_colour_bar() within Hmisc::subplot() or TeachingDemos::subplot() 
# ... to add the colour bar as a subplot 
utils::str(dat4colbar)
Hmisc::subplot(x = axis_ls[[1]]$lim[2], 
               y = axis_ls[[2]]$lim[1], 
               size = c(0.2, 2.5),
               vadj = 0, 
               hadj = 0,
  fun = add_colour_bar(data_legend = dat4colbar$data_legend, 
                       pretty_axis_args = dat4colbar$axis_legend
                       )
  )
#### Add pretty axes back to the plot 
add_pretty_axis
par(pp)

```

`add_lines()` and `add_colour_bar()` have lots of customisation options which are exemplified in the help file (use `?add_lines()` and `?add_colour_bar()`). For example, we could easily change the colour scheme by supplying a different function to create colours, or adjust the axis labels on the colour bar by passing arguments to `pretty_axis_args`. (Under-the-hood, this is calls `do.call()` to evaluate additional arguments). 

```{r fig.width = 7, fig.height = 5}
#### Define plotting window (as above)
pp <- par(oma = c(1, 1, 1, 4))

#### Plot the graph (as above)
plot2

#### Use add_lines() to add a line coloured by temperature. 
# This time, we'll adjust the number of breaks desired on the colour bar axis
# ... and the function used to colour the timeseries:
dat4colbar <- 
  add_lines(x = df$timestamp, 
            y1 = df$dn, 
            y2 = df$temp, 
            f = viridis::viridis,
            pretty_axis_args = list(pretty = list(n = 15), axis = list(las = TRUE)))

#### Use add_colour_bar() (as above)
Hmisc::subplot(x = axis_ls[[1]]$lim[2], 
               y = axis_ls[[2]]$lim[1], 
               size = c(0.2, 2.5),
               vadj = 0, 
               hadj = 0,
  fun = add_colour_bar(data_legend = dat4colbar$data_legend, 
                       pretty_axis_args = dat4colbar$axis_legend
                       )
  )
#### Add pretty axes back to the plot 
add_pretty_axis
par(pp)

```

For these sample data, this exploration suggests that the predominant axis of variation in temperature is through time, rather than through space (i.e. as the skate move shallower/deeper over short periods of time). Without any modelling, this suggests that one explanation for movements between shallow and deep water for some marine species, the so-called 'hunt-warm, rest-cold' hypothesis in which individuals are hypothesised to descend into deeper waters to maximise digestive efficiency following feeding, is probably not a suitable explanation for the vertical movements of this individual. 

## Add grids: `add_grid_rect_xy()`

To aid interpretation, it is often helpful to add a regular grid to a plot at user-specified positions. Here, we'll add a grid coinciding with pretty axis locations: 

```{r, fig.width = 7, fig.height = 5}
plot2
add_grid_rect_xy(x = axis_ls[[1]]$axis$at, y = axis_ls[[2]]$axis$at, col = scales::alpha("dimgrey", 0.8))
add_lines(x = df$timestamp, y1 = df$dn)
add_pretty_axis
```

See `?add_grid_rect_xy()` for further options. 

## Add shading: `add_shading_bar()`

Colouring lines by the value of a third variable is one useful approach in the exploration of relationships. For discrete variables, adding rectangular shading demarking switches between values is also useful for inference. For instance, in the following example, we add vertical blocks of shading delineating day/night periods:

```{r, fig.width = 7, fig.height = 5}
#### Reproduce blank plot with pretty limits
plot2

#### Define the positions of day/night periods 
# Note that Tools4ETS::define_time_blocks() could be used here to streamline the code below. 
# Define the times of day/night across the timeseries:
dates_block <- seq.POSIXt(min(df$timestamp), max(df$timestamp), by = "days")
dat_block <- data.frame(date = sort(rep(dates_block, 2)), level = factor(c(rep(1:2, length(dates_block)))))
dat_block$time <- rep(min(df$timestamp), nrow(dat_block))
coords <- matrix(c(56, -5), nrow = 1)
pos_sunrise <- which(dat_block$level == 1); pos_sunset <- which(dat_block$level == 2)
dat_block$time[pos_sunrise] <- maptools::sunriset(coords,
                                                  dateTime = dat_block$date[pos_sunrise],
                                                  direction = "sunrise",
                                                  POSIXct.out = TRUE)[,2]
dat_block$time[pos_sunset] <- maptools::sunriset(coords,
                                                 dateTime = dat_block$date[pos_sunset],
                                                 direction = "sunset",
                                                 POSIXct.out = TRUE)[,2]
# Associate day/night periods with appropriate colour
dat_block$col <- c("white", "dimgrey")[dat_block$level]
# Rearrange dataframe:
x1 <- dat_block$time[1:(nrow(dat_block)-1)]
x2 <- dat_block$time[2:nrow(dat_block)]
col <- dat_block$col[1:(nrow(dat_block)-1)]
dat_block <- data.frame(x1 = x1, x2 = x2, col = col)
dat_block$col <- as.character(dat_block$col)

#### These can also be generated by the user. All that is required is that
# ... the positions (and colours) of the blocks is specifed, here by x1, x2 and col:
utils::str(dat_block)
#### Add shading to the plot, with vertical bars exactly within pretty y axis limits 
add_shading_bar(dat_block$x1,
            dat_block$x2,
            horiz = FALSE,
            lim = axis_ls[[2]]$lim,
            col = dat_block$col, 
            border = FALSE) %>% invisible()

#### Add back depth timeseries 
add_lines(x = df$timestamp, y1 = df$dn)

#### Add pretty axes
add_pretty_axis
```

`?Tools4ETS::define_time_blocks()` and `?add_shading_bar()` provide further information about these functions. In this case, adding shading reveals that this individual often seems to ascend at night -- a common behaviour in marine organisms termed diel vertical migration -- but this behaviour is highly variable, both in extent and direction (i.e. sometimes the individual is shallower in the day). These insights from data exploration help build understanding, in this case of the movements of a poorly studied and Critically Endangered elasmobranch.   

## Add statistical summarises: `summarise_in_bins()`

Statistical summaries can also aid interpretation. `summarise_in_bins()` can be used to calculate summary statistics for continuous data in user-specified bins or breaks. With the argument `shift = TRUE`, the resulting summary statistics are shifted to the middle value of each bin, reflecting that they are a summary across that bin. With `to_plot = TRUE` additional checks are implemented to ensure the resulting summaries are retained within plot limits. For example, we can examine the change in depth in 12 hour intervals as follows:

```{r, fig.width = 7, fig.height = 5}
#### Reproduct plot 
plot1
# Summarise depth in 12 hour bins
# and define a list of summary statistics, one for each function supplied: 
summaries <- summarise_in_bins(x = df$timestamp, 
                               y = df$dn, 
                               bin = "12 hours", 
                               funs = list(mean = mean, min = min, max = max),
                               shift = TRUE,
                               to_plot = TRUE)
utils::str(summaries)

#### Add statistical summaries to the plot
# Define a sequence of colours for each summary statistic
cols <- c("red", "blue", "blue")
# Add the values of each summary statistic to the plot as a coloured line: 
mapply(summaries, cols, FUN = function(foo, col){
  add_lines(foo$bin, foo$stat, col = col, lwd = 1.5)
}) %>% invisible()

#### Add back pretty axes as usual 
add_pretty_axis
```

Further information is provided in the help file (see `?summarise_in_bins()`). Here, these simple summaries help to demonstrate that there is no clear, overall trend in the mean depth or the variation in depth through time for this individual over a two week period. Small movements seem to occur frequently within 12 hour windows, but their overall extent is relatively stable (except in the middle of the timeseries, when the individual rapidly ascended to the surface; at this point the individual was re-caught and released).

## Add model predictions: `list_CIs()` and `add_model_predictions()`

During or following data exploration, models are useful tools to elucidate trends and relationships. To evaluate the performance of models, it is often useful to plot model predictions ontop of observed data. Two functions make this easy with `prettyGraphics`: `list_CIs()` and `add_model_predictions()`. To illustrate this, below I consider a model of depth as a smooth function of time, using a thinned subset of the data, and then add predictions to a plot of the depth timeseries. Note that this model is purely for demonstration purposes; for further discussion, see the `Tools4ETS` package.

```{r, fig.width = 7, fig.height = 5}
#### Flag and thin timeseries to reduce serial autocorrelation 
dat_flapper_thin <- df[seq(1, nrow(df), by = 10), ]

#### Implement a preliminary model of depth ~ s(time) using mgcv::bam and the thinned dataset 
m1 <- mgcv::bam(depth ~ s(scale(as.numeric(timestamp), scale = FALSE), bs = "cr", k = 10), 
                rho = 0.78,
                data = dat_flapper_thin)

#### Use the predict function to define fitted values and SEs
p <- mgcv::predict.bam(m1, se.fit = TRUE)

#### Use list_CIs() to list CIs
# Since we modelled depth as a positive number, but on our plots we've been considering
# ... depth as a negative number, we'll negate fitted values by supplying a function to fadj argument 
# ... of list_CIs()
CIs <- list_CIs(pred = p, fadj = function(x){-x})
# We now have a list of fitted values, lower and upper CIs which we can plot 
utils::str(CIs)

#### Plot observed data 
plot1

#### Add model predictions using default options (i.e. CIs and fitted lines)
# Note that x values are defined for the thinned dataset, used for modelling, rather than the whole dataset 
add_model_predictions(x = dat_flapper_thin$timestamp, CI = CIs)

#### Add back pretty axes 
add_pretty_axis
```

`?list_CIs()` and `?add_model_predictions()` detail customisation options. In this case, our model reinforces our belief that the overall change in the mean depth is limited through time, although we would need to interrogate our model much more closely before putting too much confidence in its predictions (see below). 

# Integrative functions 

`pretty.plot` contains a few integrative functions which implement some of the building blocks described above for particular purposes, with more or less generality. Current functions include `pretty_residuals()`, which produces a raft of pretty diagnostic residual plots for a model, `pretty_hist()` and `pretty_density()` which produce histograms and density plots (which can be useful in data exploration and model diagnostics) and `pretty_plot()` which is very like `graphics::plot()` but uses the power of `pretty_axis()` to make the axes and can handle several function types (e.g. `stats::qqnorm()`, see below). 

## Model residuals: `pretty_residuals()`

To examine the extent to which model assumptions are met, residual diagnostic plots are instructive. These include standard diagnostic plots (i.e., residuals versus fitted values, residuals versus the linear predictor, a histogram of residuals and a quantile-quantile plot of residuals) as well as plots of residuals against covariates (`vars`, below) and, for timeseries, timestamps and autocorrelation functions (ACF) of residuals. This functionality is provided by `pretty_residuals()`, which plots standard diagnostic plots with pretty axes as well as residuals against one or more variables, timestamps and the ACF (if applicable). 

```{r, fig.width = 7, fig.height = 10}
pp <- par(mfrow = c(4, 2), oma = c(2, 2, 1, 1), mar = c(3, 3, 3, 3))
pretty_residuals(residuals = m1$std.rsd, 
               fv = stats::fitted(m1),
               lp = stats::fitted(m1), 
               vars = "temp",
               timestamp = "timestamp",
               dat = dat_flapper_thin, 
               plot = 1:7
               )
par(pp)
```

For large datasets, it can be helpful if some plots are based on random subset of residuals. For example, below we'll use all residuals for most plots (some plots, like quantile-quantile plots, can be quite sensitive to the removal of data), but only plot a random sample of 50 % of residuals for plots 3 and 4, which show the residuals versus fitted values and the linear predictor. (In this case, the two plots are identical.) 
```{r, fig.width = 7, fig.height = 10}
pp <- par(mfrow = c(4, 2), oma = c(2, 2, 1, 1), mar = c(3, 3, 3, 3))
pretty_residuals(residuals = m1$std.rsd, 
               fv = stats::fitted(m1),
               lp = stats::fitted(m1), 
               vars = "temp",
               timestamp = "timestamp",
               dat = dat_flapper_thin, 
               plot = 1:7,
               rand_pc = 50,
               plot_rand_pc = c(3, 4),
               )
par(pp)
```
More details are given in `?pretty_residuals`. For the example depth timeseries, these plots demonstrate that out model needs to be interpreted with caution: residuals versus fitted values, temperature and timestamp look reasonable, but the model has not perfectly captured the autocorrelation among residuals and the assumption of normality is clearly violated. 

## Number lines and timelines: `pretty_line()`

In some cases, one dimensional visualisation of observations, especially those collected through time, can be useful. For example, we could view a timeline of the depth timeseries observations as follows:

```{r, fig.width = 7, fig.height = 5}
pretty_line(df$timestamp, pch = 21, col = "red", bg = "red", cex = 0.1)
```

This kind of approach is particularly useful for presence or presence/absence observations collected through time. `?pretty_line()` provides further information and examples, including exemplifying the addition of one dimensional number lines or timelines to existing plots. 

## Histograms and density plots: `pretty_hist()` and `pretty_curve()`

Histograms and density plots are also useful components of data exploration and/or modelling diagnostics. Below, I use `pretty_hist()` to create a histogram of the distribution of depths and `pretty_curve()` to explore the distribution of observed data in relation to a Gaussian distribution with parameters estimated by the model above: 

```{r, fig.width = 7, fig.height = 5}
#### Plotting window 
pp <- par(mfrow = c(1, 2))

#### pretty histogram 
pretty_hist(df$depth)

#### pretty density plot 
pretty_curve(x = df$depth, 
             f = stats::dnorm, 
             param = list(mean = coef(m1)[1], sd = sigma(m1)), 
             add_rug = list(col = "skyblue", lwd = 2)
             )
par(pp)
```

## Boxplots: `pretty_boxplot()`

For factor data, boxplots are often particularly informative, and these can be implemented via `pretty_boxplot()`. For example, we could explore the distribution of depths occupied by the two sampled individuals as follows: 

```{r, fig.width = 7, fig.height = 5}
pretty_boxplot(dat_flapper$id, dat_flapper$depth, xlab = "ID", ylab = "Depth (m)")
```

## Timeseries plot: `Tools4ETS::pretty_ts()`

For timeseries data, `Tools4ETS::pretty_ts()` links most of the building blocks described above to streamline the creation of complex plots. `?Tools4ETS::pretty_ts()` provides further information and examples. `Tools4ETS::vis_ts()` is an extension of `pretty_ts()` into an interactive environment which facilitates rapid data exploration and the production of publication-quality plots for timeseries data via an R Shiny-Dashboard interface. 

## Multiple plot functions: `pretty_plot()`

`pretty_plot()` is a general function which produces plot with pretty axes. This operates like `graphics::plot()`, with a few differences:

* **Axes**. Axes are controlled by `pretty_axis()` via `pretty_axis_args`, which ensures they are always appropriately placed. 
* **Axis labels**. Labels can be added to a plot using `xlab`, `ylab` and `main`, as usual, or with `graphics::mtext()` via the `mtext_args` argument, which offers more flexibility.
* **Plot functions**. Several plotting functions can be used via the specification of a function, `f` and `plot_xy`, which describes whether the function takes in the `x` or `y` coordinates or both (for example `stats::qqnorm()` uses `y` alone, see below). 
* **Graphics**. Points and lines can be added to a plot by supplying the usual arguments (e.g. `pch`), or via named lists of arguments (`points_args` or `lines_args`) passed to `graphics::points()` or `graphics::lines()`. 

For example, we can re-create a pretty version of our depth timeseries plots, with most of the work (e.g. definition of suitable axes based on user inputs) implemented internally, via:
```{r, fig.width = 7, fig.height = 5}
pretty_plot(x = df$timestamp, 
            y = df$depth, 
            plot_xy = c("x", "y"), 
            f = graphics::plot, 
            type = "l", 
            pretty_axis_args = list(side = c(3, 2), pretty = list(n = 5), axis = list(las = TRUE)),
            mtext_args = 
              list(
                list(side = 3, text = "Time", line = 2), 
                list(side = 2, text = "Depth (m)", line = 3)
                )
            )
```

`pretty_plot()` can use some other functions, like `stats::qqnorm()`, which is implemented under-the-hood in `pretty_residuals()`, as follows:
```{r, fig.width = 7, fig.height = 5}
pp <- par(mfrow = c(1, 2))
qqnorm(m1$std.rsd)
qq <- qqnorm(m1$std.rsd, plot = FALSE)
pretty_plot(x = qq$x, 
            y = qq$y, 
            plot_xy = "y", 
            f = stats::qqnorm, 
            pretty_axis_args = list(side = 1:2, pretty = list(n = 5), axis = list(las = TRUE)), 
            mtext_args = list(
                list(side = 1, text = "Theoretical Quantiles", line = 2), 
                list(side = 2, text = "Sample Quantiles", line = 3)
                )
            )
par(pp)

```

# Conclusions 

`prettyGraphics` is an R package designed to make common data exploration tasks and plot production for continuous data easier and more flexible. The package was particularly inspired by the need to visualise and explore ecological timeseries data but most functions have wider applicability. 

`pretty_axis()` is an especially useful function in the creation of publication quality plots in base R. The subsequent addition of lines, colour bars, statistical summaries, shading and model predictions adds value to data exploration and publication-quality plots, but can be surprisingly difficult with base R. `prettyGraphics` facilitates these operations. Functions can be used as building blocks or via integrative functions, such as `pretty_plot()`. These operate similarly to functions in base R but harness some of the power of building block functions to create improved plots. 

# Future developments

`pretty.plot` has been designed especially for continuous responses and explanatory variables, although some functions (e.g. `add_shading_bar()`) help with discrete variables or factors. This functionality may be expanded in due course. Suggestions are welcome. However, the principle motivation for `prettyGraphics` remains the requirements of (ecological) research, which means that `prettyGraphics` does not aspire to provide the level of comprehensive functionality of other plotting packages (e.g. `ggplot2`, `plotly`). 

# Acknowledgements

This work was conducted during a PhD Studentship at the University of St Andrews, jointly funded by Scottish Natural Heritage and the Centre for Research into Ecological and Environmental Modelling. EL is a member of the Marine Alliance for Science and Technology Graduate School.

# Session information

```{r}
devtools::session_info()
```
